---
title: "Random Forests Challenge"
subtitle: "The Power of Weak Learners"
format:
  html: default
  pdf: default
execute:
  echo: true
  eval: true
---

# Random Forest Challenge: From Weak to Strong Learners

## The Problem: Can Many Weak Learners Beat One Strong Learner?

**Core Question:** How does the number of trees in a random forest affect predictive accuracy, and how do random forests compare to simpler approaches like linear regression?

**The Challenge:** Individual decision trees are "weak learners" with limited predictive power. Random forests combine many weak trees to create a "strong learner" that generalizes better. But how many trees do we need? Do more trees always mean better performance, or is there a point of diminishing returns?

**Our Approach:** We'll compare random forests with different numbers of trees against linear regression and individual decision trees to understand the trade-offs between complexity and performance **for this dataset**.

## Data and Methodology

We analyze the Ames Housing dataset, which contains detailed information about residential properties sold in Ames, Iowa from 2006 to 2010. This dataset is ideal for our analysis because:

- **Anticipated Non-linear Relationships:** Real estate prices have complex, non-linear relationships between features (e.g., square footage in wealthy vs. poor zip codes affects price differently)
- **Mixed Data Types:** Contains both categorical (zipCode) and numerical variables
- **Real-world Complexity:** Captures the kind of messy, real-world data where ensemble methods excel

Since we anticipate non-linear relationships, random forests are well-suited to model the relationship between features and sale price.

::: {.panel-tabset}

### R

```{r}
#| label: load-and-model-r
#| echo: true
#| message: false
#| warning: false

# Load libraries
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(randomForest))

# Load data
sales_data <- read.csv("https://raw.githubusercontent.com/flyaflya/buad442Fall2025/refs/heads/main/datasets/salesPriceData.csv")

# Prepare model data
model_data <- sales_data %>%
  select(SalePrice, LotArea, YearBuilt, GrLivArea, FullBath, HalfBath, 
         BedroomAbvGr, TotRmsAbvGrd, GarageCars, zipCode) %>%
  # Convert zipCode to factor (categorical variable) - important for proper modeling
  mutate(zipCode = as.factor(zipCode)) %>%
  na.omit()

cat("Data prepared with zipCode as categorical variable\n")
cat("Number of unique zip codes:", length(unique(model_data$zipCode)), "\n")

# Split data
set.seed(123)
train_indices <- sample(1:nrow(model_data), 0.8 * nrow(model_data))
train_data <- model_data[train_indices, ]
test_data <- model_data[-train_indices, ]

# Build random forests with different numbers of trees (with corrected categorical zipCode)
rf_1 <- randomForest(SalePrice ~ ., data = train_data, ntree = 1, mtry = 3, seed = 123)
rf_5 <- randomForest(SalePrice ~ ., data = train_data, ntree = 5, mtry = 3, seed = 123)
rf_25 <- randomForest(SalePrice ~ ., data = train_data, ntree = 25, mtry = 3, seed = 123)
rf_100 <- randomForest(SalePrice ~ ., data = train_data, ntree = 100, mtry = 3, seed = 123)
rf_500 <- randomForest(SalePrice ~ ., data = train_data, ntree = 500, mtry = 3, seed = 123)
rf_1000 <- randomForest(SalePrice ~ ., data = train_data, ntree = 1000, mtry = 3, seed = 123)
rf_2000 <- randomForest(SalePrice ~ ., data = train_data, ntree = 2000, mtry = 3, seed = 123)
rf_5000 <- randomForest(SalePrice ~ ., data = train_data, ntree = 5000, mtry = 3, seed = 123)
```

### Python

```{python}
#| label: load-and-model-python
#| echo: true

import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from sklearn.ensemble import RandomForestRegressor
from sklearn.model_selection import train_test_split
from sklearn.metrics import mean_squared_error, r2_score
import warnings
warnings.filterwarnings('ignore')

# Load data
sales_data = pd.read_csv("https://raw.githubusercontent.com/flyaflya/buad442Fall2025/refs/heads/main/datasets/salesPriceData.csv")

# Prepare model data
model_vars = ['SalePrice', 'LotArea', 'YearBuilt', 'GrLivArea', 'FullBath', 
              'HalfBath', 'BedroomAbvGr', 'TotRmsAbvGrd', 'GarageCars', 'zipCode']
model_data = sales_data[model_vars].dropna()

# Convert zipCode to categorical variable - important for proper modeling
model_data['zipCode'] = model_data['zipCode'].astype('category')

# Split data
X = model_data.drop('SalePrice', axis=1)
y = model_data['SalePrice']
X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=0.2, random_state=123)

# Build random forests with different numbers of trees (with corrected categorical zipCode)
rf_1 = RandomForestRegressor(n_estimators=1, max_features=3, random_state=123)
rf_5 = RandomForestRegressor(n_estimators=5, max_features=3, random_state=123)
rf_25 = RandomForestRegressor(n_estimators=25, max_features=3, random_state=123)
rf_100 = RandomForestRegressor(n_estimators=100, max_features=3, random_state=123)
rf_500 = RandomForestRegressor(n_estimators=500, max_features=3, random_state=123)
rf_1000 = RandomForestRegressor(n_estimators=1000, max_features=3, random_state=123)
rf_2000 = RandomForestRegressor(n_estimators=2000, max_features=3, random_state=123)
rf_5000 = RandomForestRegressor(n_estimators=5000, max_features=3, random_state=123)

# Fit all models
rf_1.fit(X_train, y_train)
rf_5.fit(X_train, y_train)
rf_25.fit(X_train, y_train)
rf_100.fit(X_train, y_train)
rf_500.fit(X_train, y_train)
rf_1000.fit(X_train, y_train)
rf_2000.fit(X_train, y_train)
rf_5000.fit(X_train, y_train)
```

:::

## Results: The Power of Ensemble Learning

Our analysis reveals a clear pattern: **more trees consistently improve performance**. Let's examine the results and understand why this happens.

### Performance Trends

::: {.panel-tabset}

### R

```{r}
#| label: performance-comparison-r
#| echo: true
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Calculate predictions and performance metrics for test data
predictions_1_test <- predict(rf_1, test_data)
predictions_5_test <- predict(rf_5, test_data)
predictions_25_test <- predict(rf_25, test_data)
predictions_100_test <- predict(rf_100, test_data)
predictions_500_test <- predict(rf_500, test_data)
predictions_1000_test <- predict(rf_1000, test_data)
predictions_2000_test <- predict(rf_2000, test_data)
predictions_5000_test <- predict(rf_5000, test_data)

# Calculate predictions for training data
predictions_1_train <- predict(rf_1, train_data)
predictions_5_train <- predict(rf_5, train_data)
predictions_25_train <- predict(rf_25, train_data)
predictions_100_train <- predict(rf_100, train_data)
predictions_500_train <- predict(rf_500, train_data)
predictions_1000_train <- predict(rf_1000, train_data)
predictions_2000_train <- predict(rf_2000, train_data)
predictions_5000_train <- predict(rf_5000, train_data)

# Calculate RMSE for test data
rmse_1_test <- sqrt(mean((test_data$SalePrice - predictions_1_test)^2))
rmse_5_test <- sqrt(mean((test_data$SalePrice - predictions_5_test)^2))
rmse_25_test <- sqrt(mean((test_data$SalePrice - predictions_25_test)^2))
rmse_100_test <- sqrt(mean((test_data$SalePrice - predictions_100_test)^2))
rmse_500_test <- sqrt(mean((test_data$SalePrice - predictions_500_test)^2))
rmse_1000_test <- sqrt(mean((test_data$SalePrice - predictions_1000_test)^2))
rmse_2000_test <- sqrt(mean((test_data$SalePrice - predictions_2000_test)^2))
rmse_5000_test <- sqrt(mean((test_data$SalePrice - predictions_5000_test)^2))

# Calculate RMSE for training data
rmse_1_train <- sqrt(mean((train_data$SalePrice - predictions_1_train)^2))
rmse_5_train <- sqrt(mean((train_data$SalePrice - predictions_5_train)^2))
rmse_25_train <- sqrt(mean((train_data$SalePrice - predictions_25_train)^2))
rmse_100_train <- sqrt(mean((train_data$SalePrice - predictions_100_train)^2))
rmse_500_train <- sqrt(mean((train_data$SalePrice - predictions_500_train)^2))
rmse_1000_train <- sqrt(mean((train_data$SalePrice - predictions_1000_train)^2))
rmse_2000_train <- sqrt(mean((train_data$SalePrice - predictions_2000_train)^2))
rmse_5000_train <- sqrt(mean((train_data$SalePrice - predictions_5000_train)^2))

# Calculate R-squared
r2_1 <- 1 - sum((test_data$SalePrice - predictions_1_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_5 <- 1 - sum((test_data$SalePrice - predictions_5_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_25 <- 1 - sum((test_data$SalePrice - predictions_25_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_100 <- 1 - sum((test_data$SalePrice - predictions_100_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_500 <- 1 - sum((test_data$SalePrice - predictions_500_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_1000 <- 1 - sum((test_data$SalePrice - predictions_1000_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_2000 <- 1 - sum((test_data$SalePrice - predictions_2000_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_5000 <- 1 - sum((test_data$SalePrice - predictions_5000_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)

# Create performance comparison
performance_df <- data.frame(
  Trees = c(1, 5, 25, 100, 500, 1000, 2000, 5000),
  RMSE_Test = c(rmse_1_test, rmse_5_test, rmse_25_test, rmse_100_test, rmse_500_test, rmse_1000_test, rmse_2000_test, rmse_5000_test),
  RMSE_Train = c(rmse_1_train, rmse_5_train, rmse_25_train, rmse_100_train, rmse_500_train, rmse_1000_train, rmse_2000_train, rmse_5000_train),
  R_squared = c(r2_1, r2_5, r2_25, r2_100, r2_500, r2_1000, r2_2000, r2_5000)
)

print(performance_df)
```

### Python

```{python}
#| label: performance-comparison-python
#| echo: true
#| fig-width: 10
#| fig-height: 6

# Calculate predictions for test data
predictions_1_test = rf_1.predict(X_test)
predictions_5_test = rf_5.predict(X_test)
predictions_25_test = rf_25.predict(X_test)
predictions_100_test = rf_100.predict(X_test)
predictions_500_test = rf_500.predict(X_test)
predictions_1000_test = rf_1000.predict(X_test)
predictions_2000_test = rf_2000.predict(X_test)
predictions_5000_test = rf_5000.predict(X_test)

# Calculate predictions for training data
predictions_1_train = rf_1.predict(X_train)
predictions_5_train = rf_5.predict(X_train)
predictions_25_train = rf_25.predict(X_train)
predictions_100_train = rf_100.predict(X_train)
predictions_500_train = rf_500.predict(X_train)
predictions_1000_train = rf_1000.predict(X_train)
predictions_2000_train = rf_2000.predict(X_train)
predictions_5000_train = rf_5000.predict(X_train)

# Calculate performance metrics for test data
rmse_1_test = np.sqrt(mean_squared_error(y_test, predictions_1_test))
rmse_5_test = np.sqrt(mean_squared_error(y_test, predictions_5_test))
rmse_25_test = np.sqrt(mean_squared_error(y_test, predictions_25_test))
rmse_100_test = np.sqrt(mean_squared_error(y_test, predictions_100_test))
rmse_500_test = np.sqrt(mean_squared_error(y_test, predictions_500_test))
rmse_1000_test = np.sqrt(mean_squared_error(y_test, predictions_1000_test))
rmse_2000_test = np.sqrt(mean_squared_error(y_test, predictions_2000_test))
rmse_5000_test = np.sqrt(mean_squared_error(y_test, predictions_5000_test))

# Calculate performance metrics for training data
rmse_1_train = np.sqrt(mean_squared_error(y_train, predictions_1_train))
rmse_5_train = np.sqrt(mean_squared_error(y_train, predictions_5_train))
rmse_25_train = np.sqrt(mean_squared_error(y_train, predictions_25_train))
rmse_100_train = np.sqrt(mean_squared_error(y_train, predictions_100_train))
rmse_500_train = np.sqrt(mean_squared_error(y_train, predictions_500_train))
rmse_1000_train = np.sqrt(mean_squared_error(y_train, predictions_1000_train))
rmse_2000_train = np.sqrt(mean_squared_error(y_train, predictions_2000_train))
rmse_5000_train = np.sqrt(mean_squared_error(y_train, predictions_5000_train))

r2_1 = r2_score(y_test, predictions_1_test)
r2_5 = r2_score(y_test, predictions_5_test)
r2_25 = r2_score(y_test, predictions_25_test)
r2_100 = r2_score(y_test, predictions_100_test)
r2_500 = r2_score(y_test, predictions_500_test)
r2_1000 = r2_score(y_test, predictions_1000_test)
r2_2000 = r2_score(y_test, predictions_2000_test)
r2_5000 = r2_score(y_test, predictions_5000_test)

# Create performance comparison
performance_data = {
    'Trees': [1, 5, 25, 100, 500, 1000, 2000, 5000],
    'RMSE_Test': [rmse_1_test, rmse_5_test, rmse_25_test, rmse_100_test, rmse_500_test, rmse_1000_test, rmse_2000_test, rmse_5000_test],
    'RMSE_Train': [rmse_1_train, rmse_5_train, rmse_25_train, rmse_100_train, rmse_500_train, rmse_1000_train, rmse_2000_train, rmse_5000_train],
    'R_squared': [r2_1, r2_5, r2_25, r2_100, r2_500, r2_1000, r2_2000, r2_5000]
}

performance_df = pd.DataFrame(performance_data)
print(performance_df)
```

:::

## Performance Visualization: The Power of More Trees

::: {.panel-tabset}

### R

```{r}
#| label: performance-plot-r
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Create performance visualization
library(ggplot2)

# Plot RMSE vs Number of Trees
p1 <- ggplot(performance_df, aes(x = Trees, y = RMSE_Test)) +
  geom_line(color = "red", size = 1.2) +
  geom_point(color = "red", size = 3) +
  labs(title = "RMSE Decreases with More Trees",
       x = "Number of Trees", y = "RMSE") +
  theme_minimal() +
  scale_x_log10()

# Plot R-squared vs Number of Trees
p2 <- ggplot(performance_df, aes(x = Trees, y = R_squared)) +
  geom_line(color = "blue", size = 1.2) +
  geom_point(color = "blue", size = 3) +
  labs(title = "R-squared Increases with More Trees",
       x = "Number of Trees", y = "R-squared") +
  theme_minimal() +
  scale_x_log10()

# Combine plots
library(gridExtra)
grid.arrange(p1, p2, ncol = 2)
```

### Python

```{python}
#| label: performance-plot-python
#| echo: false
#| fig-width: 10
#| fig-height: 6

# Create performance visualization
fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 5))

# Plot RMSE vs Number of Trees
ax1.plot(performance_df['Trees'], performance_df['RMSE_Test'], 'ro-', linewidth=2, markersize=8)
ax1.set_xlabel('Number of Trees')
ax1.set_ylabel('RMSE')
ax1.set_title('RMSE Decreases with More Trees')
ax1.set_xscale('log')
ax1.grid(True, alpha=0.3)

# Plot R-squared vs Number of Trees
ax2.plot(performance_df['Trees'], performance_df['R_squared'], 'bo-', linewidth=2, markersize=8)
ax2.set_xlabel('Number of Trees')
ax2.set_ylabel('R-squared')
ax2.set_title('R-squared Increases with More Trees')
ax2.set_xscale('log')
ax2.grid(True, alpha=0.3)

plt.tight_layout()
plt.show()
```

:::

The results show a clear trend: as we increase the number of trees from 1 to 5,000, both RMSE decreases and R-squared increases. This demonstrates the fundamental principle of ensemble learning analogous to the 'wisdom of crowds' phenomenon.

### Important Notes on Random Forest Performance

**The Ensemble Effect:** Individual decision trees make many errors, but when combined, their errors cancel out. This is why random forests with more trees consistently outperform those with fewer trees.

**No Overfitting:** Unlike other algorithms, adding more trees to a random forest doesn't cause overfitting - it only improves performance.

**Diminishing Returns:** While more trees always help, the improvement becomes smaller as we add more trees.

## The Overfitting Problem: Why Random Forests Are Different

To understand why random forests are so powerful, let's compare them with individual decision trees of increasing complexity.

### The Problem with Single Trees

As we make individual trees more complex (deeper, more splits), they start to memorize the training data and perform worse on new data. This is classic overfitting.

::: {.panel-tabset}

### R

```{r}
#| label: decision-tree-comparison-r
#| echo: true
#| message: false
#| warning: false

# Build decision trees with increasing complexity (with corrected categorical zipCode)
library(rpart)
tree_simple <- rpart(SalePrice ~ ., data = train_data, 
                     control = rpart.control(maxdepth = 2, minsplit = 20, minbucket = 10))
tree_medium <- rpart(SalePrice ~ ., data = train_data, 
                     control = rpart.control(maxdepth = 4, minsplit = 20, minbucket = 10))
tree_complex <- rpart(SalePrice ~ ., data = train_data, 
                     control = rpart.control(maxdepth = 6, minsplit = 15, minbucket = 8))
tree_very_complex <- rpart(SalePrice ~ ., data = train_data, 
                          control = rpart.control(maxdepth = 8, minsplit = 10, minbucket = 5))
tree_extreme <- rpart(SalePrice ~ ., data = train_data, 
                     control = rpart.control(maxdepth = 12, minsplit = 5, minbucket = 2))
tree_overfit <- rpart(SalePrice ~ ., data = train_data, 
                     control = rpart.control(maxdepth = 20, minsplit = 2, minbucket = 1))

# Calculate predictions for test data
pred_simple_test <- predict(tree_simple, test_data)
pred_medium_test <- predict(tree_medium, test_data)
pred_complex_test <- predict(tree_complex, test_data)
pred_very_complex_test <- predict(tree_very_complex, test_data)
pred_extreme_test <- predict(tree_extreme, test_data)
pred_overfit_test <- predict(tree_overfit, test_data)

# Calculate predictions for training data
pred_simple_train <- predict(tree_simple, train_data)
pred_medium_train <- predict(tree_medium, train_data)
pred_complex_train <- predict(tree_complex, train_data)
pred_very_complex_train <- predict(tree_very_complex, train_data)
pred_extreme_train <- predict(tree_extreme, train_data)
pred_overfit_train <- predict(tree_overfit, train_data)

# Calculate RMSE for test data
rmse_tree_simple_test <- sqrt(mean((test_data$SalePrice - pred_simple_test)^2))
rmse_tree_medium_test <- sqrt(mean((test_data$SalePrice - pred_medium_test)^2))
rmse_tree_complex_test <- sqrt(mean((test_data$SalePrice - pred_complex_test)^2))
rmse_tree_very_complex_test <- sqrt(mean((test_data$SalePrice - pred_very_complex_test)^2))
rmse_tree_extreme_test <- sqrt(mean((test_data$SalePrice - pred_extreme_test)^2))
rmse_tree_overfit_test <- sqrt(mean((test_data$SalePrice - pred_overfit_test)^2))

# Calculate RMSE for training data
rmse_tree_simple_train <- sqrt(mean((train_data$SalePrice - pred_simple_train)^2))
rmse_tree_medium_train <- sqrt(mean((train_data$SalePrice - pred_medium_train)^2))
rmse_tree_complex_train <- sqrt(mean((train_data$SalePrice - pred_complex_train)^2))
rmse_tree_very_complex_train <- sqrt(mean((train_data$SalePrice - pred_very_complex_train)^2))
rmse_tree_extreme_train <- sqrt(mean((train_data$SalePrice - pred_extreme_train)^2))
rmse_tree_overfit_train <- sqrt(mean((train_data$SalePrice - pred_overfit_train)^2))

# Calculate R-squared for test data
r2_tree_simple <- 1 - sum((test_data$SalePrice - pred_simple_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_tree_medium <- 1 - sum((test_data$SalePrice - pred_medium_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_tree_complex <- 1 - sum((test_data$SalePrice - pred_complex_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_tree_very_complex <- 1 - sum((test_data$SalePrice - pred_very_complex_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_tree_extreme <- 1 - sum((test_data$SalePrice - pred_extreme_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_tree_overfit <- 1 - sum((test_data$SalePrice - pred_overfit_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)

# Create comparison data frame
tree_comparison <- data.frame(
  Model = c("Simple Tree", "Medium Tree", "Complex Tree", "Very Complex Tree", "Extreme Tree", "Overfit Tree"),
  Max_Depth = c(2, 4, 6, 8, 12, 20),
  RMSE_Test = c(rmse_tree_simple_test, rmse_tree_medium_test, rmse_tree_complex_test, rmse_tree_very_complex_test, rmse_tree_extreme_test, rmse_tree_overfit_test),
  RMSE_Train = c(rmse_tree_simple_train, rmse_tree_medium_train, rmse_tree_complex_train, rmse_tree_very_complex_train, rmse_tree_extreme_train, rmse_tree_overfit_train),
  R_squared = c(r2_tree_simple, r2_tree_medium, r2_tree_complex, r2_tree_very_complex, r2_tree_extreme, r2_tree_overfit)
)

print("Decision Tree Performance (showing overfitting):")
print(tree_comparison)
```

### Python

```{python}
#| label: decision-tree-comparison-python
#| echo: true

# Build decision trees with increasing complexity (with corrected categorical zipCode)
from sklearn.tree import DecisionTreeRegressor

tree_simple = DecisionTreeRegressor(max_depth=2, min_samples_split=20, min_samples_leaf=10, random_state=123)
tree_medium = DecisionTreeRegressor(max_depth=4, min_samples_split=20, min_samples_leaf=10, random_state=123)
tree_complex = DecisionTreeRegressor(max_depth=6, min_samples_split=15, min_samples_leaf=8, random_state=123)
tree_very_complex = DecisionTreeRegressor(max_depth=8, min_samples_split=10, min_samples_leaf=5, random_state=123)
tree_extreme = DecisionTreeRegressor(max_depth=12, min_samples_split=5, min_samples_leaf=2, random_state=123)
tree_overfit = DecisionTreeRegressor(max_depth=20, min_samples_split=2, min_samples_leaf=1, random_state=123)

# Fit models
tree_simple.fit(X_train, y_train)
tree_medium.fit(X_train, y_train)
tree_complex.fit(X_train, y_train)
tree_very_complex.fit(X_train, y_train)
tree_extreme.fit(X_train, y_train)
tree_overfit.fit(X_train, y_train)

# Calculate predictions for test data
pred_simple_test = tree_simple.predict(X_test)
pred_medium_test = tree_medium.predict(X_test)
pred_complex_test = tree_complex.predict(X_test)
pred_very_complex_test = tree_very_complex.predict(X_test)
pred_extreme_test = tree_extreme.predict(X_test)
pred_overfit_test = tree_overfit.predict(X_test)

# Calculate predictions for training data
pred_simple_train = tree_simple.predict(X_train)
pred_medium_train = tree_medium.predict(X_train)
pred_complex_train = tree_complex.predict(X_train)
pred_very_complex_train = tree_very_complex.predict(X_train)
pred_extreme_train = tree_extreme.predict(X_train)
pred_overfit_train = tree_overfit.predict(X_train)

# Calculate performance metrics for test data
rmse_tree_simple_test = np.sqrt(mean_squared_error(y_test, pred_simple_test))
rmse_tree_medium_test = np.sqrt(mean_squared_error(y_test, pred_medium_test))
rmse_tree_complex_test = np.sqrt(mean_squared_error(y_test, pred_complex_test))
rmse_tree_very_complex_test = np.sqrt(mean_squared_error(y_test, pred_very_complex_test))
rmse_tree_extreme_test = np.sqrt(mean_squared_error(y_test, pred_extreme_test))
rmse_tree_overfit_test = np.sqrt(mean_squared_error(y_test, pred_overfit_test))

# Calculate performance metrics for training data
rmse_tree_simple_train = np.sqrt(mean_squared_error(y_train, pred_simple_train))
rmse_tree_medium_train = np.sqrt(mean_squared_error(y_train, pred_medium_train))
rmse_tree_complex_train = np.sqrt(mean_squared_error(y_train, pred_complex_train))
rmse_tree_very_complex_train = np.sqrt(mean_squared_error(y_train, pred_very_complex_train))
rmse_tree_extreme_train = np.sqrt(mean_squared_error(y_train, pred_extreme_train))
rmse_tree_overfit_train = np.sqrt(mean_squared_error(y_train, pred_overfit_train))

r2_tree_simple = r2_score(y_test, pred_simple_test)
r2_tree_medium = r2_score(y_test, pred_medium_test)
r2_tree_complex = r2_score(y_test, pred_complex_test)
r2_tree_very_complex = r2_score(y_test, pred_very_complex_test)
r2_tree_extreme = r2_score(y_test, pred_extreme_test)
r2_tree_overfit = r2_score(y_test, pred_overfit_test)

# Create comparison data frame
tree_comparison_data = {
    'Model': ['Simple Tree', 'Medium Tree', 'Complex Tree', 'Very Complex Tree', 'Extreme Tree', 'Overfit Tree'],
    'Max_Depth': [2, 4, 6, 8, 12, 20],
    'RMSE_Test': [rmse_tree_simple_test, rmse_tree_medium_test, rmse_tree_complex_test, rmse_tree_very_complex_test, rmse_tree_extreme_test, rmse_tree_overfit_test],
    'RMSE_Train': [rmse_tree_simple_train, rmse_tree_medium_train, rmse_tree_complex_train, rmse_tree_very_complex_train, rmse_tree_extreme_train, rmse_tree_overfit_train],
    'R_squared': [r2_tree_simple, r2_tree_medium, r2_tree_complex, r2_tree_very_complex, r2_tree_extreme, r2_tree_overfit]
}

tree_comparison_df = pd.DataFrame(tree_comparison_data)
print("Decision Tree Performance (showing overfitting):")
print(tree_comparison_df)
```

:::

### The Overfitting Visualization

::: {.panel-tabset}

### R

```{r}
#| label: overfitting-plot-r
#| echo: false
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 8

# Create overfitting comparison plot
library(ggplot2)
library(gridExtra)

# Calculate common y-axis limits for both plots
all_rmse_values <- c(tree_comparison$RMSE_Test, tree_comparison$RMSE_Train, performance_df$RMSE_Test, performance_df$RMSE_Train)
y_min <- min(all_rmse_values) * 0.95
y_max <- max(all_rmse_values) * 1.05

# Plot 1: Decision Trees (showing overfitting)
p1 <- ggplot(tree_comparison, aes(x = Max_Depth)) +
  geom_line(aes(y = RMSE_Test, color = "Test Data"), size = 1.2) +
  geom_point(aes(y = RMSE_Test, color = "Test Data"), size = 3) +
  geom_line(aes(y = RMSE_Train, color = "Training Data"), size = 1.2) +
  geom_point(aes(y = RMSE_Train, color = "Training Data"), size = 3) +
  labs(title = "Decision Trees: Overfitting Problem",
       subtitle = "Training performance improves, test performance gets worse",
       x = "Max Depth", y = "RMSE", color = "Data Type") +
  theme_minimal() +
  ylim(y_min, y_max) +
  scale_color_manual(values = c("Test Data" = "red", "Training Data" = "darkred"))

# Plot 2: Random Forests (no overfitting)
p2 <- ggplot(performance_df, aes(x = Trees)) +
  geom_line(aes(y = RMSE_Test, color = "Test Data"), size = 1.2) +
  geom_point(aes(y = RMSE_Test, color = "Test Data"), size = 3) +
  geom_line(aes(y = RMSE_Train, color = "Training Data"), size = 1.2) +
  geom_point(aes(y = RMSE_Train, color = "Training Data"), size = 3) +
  labs(title = "Random Forests: No Overfitting",
       subtitle = "Both training and test performance improve with more trees",
       x = "Number of Trees", y = "RMSE", color = "Data Type") +
  theme_minimal() +
  scale_x_log10() +
  ylim(y_min, y_max) +
  scale_color_manual(values = c("Test Data" = "blue", "Training Data" = "darkblue"))

# Combine plots
grid.arrange(p1, p2, ncol = 2)
```

### Python

```{python}
#| label: overfitting-plot-python
#| echo: false
#| fig-width: 12
#| fig-height: 8

# Create overfitting comparison plot
fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(15, 6))

# Calculate common y-axis limits for both plots
all_rmse_values = list(tree_comparison_df['RMSE_Test']) + list(tree_comparison_df['RMSE_Train']) + list(performance_df['RMSE_Test']) + list(performance_df['RMSE_Train'])
y_min = min(all_rmse_values) * 0.95
y_max = max(all_rmse_values) * 1.05

# Plot 1: Decision Trees (showing overfitting)
ax1.plot(tree_comparison_df['Max_Depth'], tree_comparison_df['RMSE_Test'], 'ro-', linewidth=2, markersize=8, label='Test Data')
ax1.plot(tree_comparison_df['Max_Depth'], tree_comparison_df['RMSE_Train'], 'darkred', linestyle='--', linewidth=2, markersize=8, label='Training Data')
ax1.set_xlabel('Max Depth')
ax1.set_ylabel('RMSE')
ax1.set_title('Decision Trees: Overfitting Problem\nTraining improves, test gets worse')
ax1.set_ylim(y_min, y_max)
ax1.legend()
ax1.grid(True, alpha=0.3)

# Plot 2: Random Forests (no overfitting)
ax2.plot(performance_df['Trees'], performance_df['RMSE_Test'], 'bo-', linewidth=2, markersize=8, label='Test Data')
ax2.plot(performance_df['Trees'], performance_df['RMSE_Train'], 'darkblue', linestyle='--', linewidth=2, markersize=8, label='Training Data')
ax2.set_xlabel('Number of Trees')
ax2.set_ylabel('RMSE')
ax2.set_title('Random Forests: No Overfitting\nBoth training and test improve with more trees')
ax2.set_xscale('log')
ax2.set_ylim(y_min, y_max)
ax2.legend()
ax2.grid(True, alpha=0.3)

plt.tight_layout()
plt.show()
```

:::

### The Random Forest Solution

**Why Random Forests Don't Overfit:**
1. **Bootstrap Sampling:** Each tree sees different data, preventing memorization
2. **Random Feature Selection:** Each tree uses different features, increasing diversity  
3. **Averaging Effect:** Multiple diverse predictions cancel out individual errors
4. **No Single Tree Dominates:** No one tree can overfit the entire dataset

**The Magic:** Random forests tend to not overfit and provide a robust prediction.

## Baseline Comparison: Random Forest vs Linear Regression

To put random forest performance in context, let's compare it to a simple linear regression baseline. This helps us understand the value of the ensemble approach.

**Why This Comparison Matters:** Linear regression provides a simple, interpretable baseline. If random forests can't significantly outperform this simple approach, their complexity may not be justified **for this dataset**.

## Linear Regression Models

::: {.panel-tabset}

### R

```{r}
#| label: linear-regression-r
#| echo: true
#| message: false
#| warning: false

# Build linear regression model
lm_model <- lm(SalePrice ~ ., data = train_data)

# Calculate predictions
lm_pred_test <- predict(lm_model, test_data)
lm_pred_train <- predict(lm_model, train_data)

# Calculate performance metrics
lm_rmse_test <- sqrt(mean((test_data$SalePrice - lm_pred_test)^2))
lm_rmse_train <- sqrt(mean((train_data$SalePrice - lm_pred_train)^2))
lm_r2 <- 1 - sum((test_data$SalePrice - lm_pred_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)

cat("Linear Regression Performance:\n")
cat("Test RMSE:", round(lm_rmse_test, 2), "\n")
cat("Train RMSE:", round(lm_rmse_train, 2), "\n")
cat("R-squared:", round(lm_r2, 4), "\n")
```

### Python

```{python}
#| label: linear-regression-python
#| echo: true

# Build linear regression model
from sklearn.linear_model import LinearRegression

lm_model = LinearRegression()
lm_model.fit(X_train, y_train)

# Calculate predictions
lm_pred_test = lm_model.predict(X_test)
lm_pred_train = lm_model.predict(X_train)

# Calculate performance metrics
lm_rmse_test = np.sqrt(mean_squared_error(y_test, lm_pred_test))
lm_rmse_train = np.sqrt(mean_squared_error(y_train, lm_pred_train))
lm_r2 = r2_score(y_test, lm_pred_test)

print("Linear Regression Performance:")
print(f"Test RMSE: {lm_rmse_test:.2f}")
print(f"Train RMSE: {lm_rmse_train:.2f}")
print(f"R-squared: {lm_r2:.4f}")
```

:::

## Performance Comparison: Random Forest vs Linear Regression

::: {.panel-tabset}

### R

```{r}
#| label: comparison-table-r
#| echo: true
#| message: false
#| warning: false

# Create comparison table (using performance_df from main analysis)
comparison_data <- data.frame(
  Model = c("Linear Regression", "Random Forest (1 tree)", "Random Forest (100 trees)", "Random Forest (1000 trees)", "Random Forest (5000 trees)"),
  Test_RMSE = c(lm_rmse_test, performance_df$RMSE_Test[1], performance_df$RMSE_Test[4], performance_df$RMSE_Test[6], performance_df$RMSE_Test[8]),
  Train_RMSE = c(lm_rmse_train, performance_df$RMSE_Train[1], performance_df$RMSE_Train[4], performance_df$RMSE_Train[6], performance_df$RMSE_Train[8]),
  R_squared = c(lm_r2, performance_df$R_squared[1], performance_df$R_squared[4], performance_df$R_squared[6], performance_df$R_squared[8]),
  Improvement_vs_LR = c(0, round((lm_rmse_test - performance_df$RMSE_Test[1])/lm_rmse_test * 100, 1), 
                       round((lm_rmse_test - performance_df$RMSE_Test[4])/lm_rmse_test * 100, 1),
                       round((lm_rmse_test - performance_df$RMSE_Test[6])/lm_rmse_test * 100, 1),
                       round((lm_rmse_test - performance_df$RMSE_Test[8])/lm_rmse_test * 100, 1))
)

print("Performance Comparison:")
print(comparison_data)
```

### Python

```{python}
#| label: comparison-table-python
#| echo: true

# Create comparison table (using performance_df from main analysis)
comparison_data = {
    'Model': ['Linear Regression', 'Random Forest (1 tree)', 'Random Forest (100 trees)', 'Random Forest (1000 trees)', 'Random Forest (5000 trees)'],
    'Test_RMSE': [lm_rmse_test, performance_df['RMSE_Test'].iloc[0], performance_df['RMSE_Test'].iloc[3], performance_df['RMSE_Test'].iloc[5], performance_df['RMSE_Test'].iloc[7]],
    'Train_RMSE': [lm_rmse_train, performance_df['RMSE_Train'].iloc[0], performance_df['RMSE_Train'].iloc[3], performance_df['RMSE_Train'].iloc[5], performance_df['RMSE_Train'].iloc[7]],
    'R_squared': [lm_r2, performance_df['R_squared'].iloc[0], performance_df['R_squared'].iloc[3], performance_df['R_squared'].iloc[5], performance_df['R_squared'].iloc[7]],
    'Improvement_vs_LR': [0, round((lm_rmse_test - performance_df['RMSE_Test'].iloc[0])/lm_rmse_test * 100, 1), 
                         round((lm_rmse_test - performance_df['RMSE_Test'].iloc[3])/lm_rmse_test * 100, 1),
                         round((lm_rmse_test - performance_df['RMSE_Test'].iloc[5])/lm_rmse_test * 100, 1),
                         round((lm_rmse_test - performance_df['RMSE_Test'].iloc[7])/lm_rmse_test * 100, 1)]
}

comparison_df = pd.DataFrame(comparison_data)
print("Performance Comparison:")
print(comparison_df)
```

:::

## Visual Comparison: RMSE Performance

::: {.panel-tabset}

### R

```{r}
#| label: comparison-plot-r
#| echo: false
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 8

# Create comparison visualization
library(ggplot2)
library(gridExtra)

# Plot 1: RMSE Comparison
p1 <- ggplot(comparison_data, aes(x = Model, y = Test_RMSE)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_text(aes(label = round(Test_RMSE, 0)), vjust = -0.5, size = 3) +
  labs(title = "Test RMSE Comparison: Linear Regression vs Random Forests",
       x = "Model", y = "Test RMSE") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Plot 2: Improvement over Linear Regression
improvement_data <- comparison_data[2:5, ] # Exclude linear regression itself
p2 <- ggplot(improvement_data, aes(x = Model, y = Improvement_vs_LR)) +
  geom_col(fill = "darkgreen", alpha = 0.7) +
  geom_text(aes(label = paste0(Improvement_vs_LR, "%")), vjust = -0.5, size = 3) +
  labs(title = "Improvement over Linear Regression",
       x = "Random Forest Model", y = "RMSE Improvement (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Combine plots
grid.arrange(p1, p2, ncol = 2)
```

### Python

```{python}
#| label: comparison-plot-python
#| echo: false
#| fig-width: 12
#| fig-height: 8

# Create comparison visualization
fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(15, 6))

# Plot 1: RMSE Comparison
ax1.bar(comparison_df['Model'], comparison_df['Test_RMSE'], color='steelblue', alpha=0.7)
ax1.set_title('Test RMSE Comparison: Linear Regression vs Random Forests')
ax1.set_xlabel('Model')
ax1.set_ylabel('Test RMSE')
ax1.tick_params(axis='x', rotation=45)

# Add value labels on bars
for i, v in enumerate(comparison_df['Test_RMSE']):
    ax1.text(i, v + 50, f'{v:.0f}', ha='center', va='bottom')

# Plot 2: Improvement over Linear Regression
improvement_df = comparison_df[1:5]  # Exclude linear regression itself
ax2.bar(improvement_df['Model'], improvement_df['Improvement_vs_LR'], color='darkgreen', alpha=0.7)
ax2.set_title('Improvement over Linear Regression')
ax2.set_xlabel('Random Forest Model')
ax2.set_ylabel('RMSE Improvement (%)')
ax2.tick_params(axis='x', rotation=45)

# Add value labels on bars
for i, v in enumerate(improvement_df['Improvement_vs_LR']):
    ax2.text(i, v + 0.5, f'{v:.1f}%', ha='center', va='bottom')

plt.tight_layout()
plt.show()
```

:::

## Key Insights: When Random Forests Excel and When They Don't

**Random Forest Advantages:**
1. **Non-linear Relationships:** Captures complex interactions between features that linear regression misses
2. **Feature Interactions:** Automatically discovers relationships between variables (e.g., zip code × lot area)
3. **Robustness:** Less sensitive to outliers and non-normal distributions
4. **Ensemble Effect:** Multiple trees reduce prediction variance

## Conclusion: The Right Tool for the Right Job

Our analysis reveals an important lesson: **random forests didn't significantly outperform linear regression on this dataset**. This outcome, while initially surprising, provides valuable insights:

**Why Random Forests Underperformed Here:**
- **Homogeneous Dataset:** The Ames housing data comes from a single, relatively uniform market (Ames, Iowa)
- **Limited Complexity:** The relationships between features and house prices may be more linear than anticipated
- **Simple Baseline:** Linear regression provides a strong baseline for this particular dataset

**When to Use Random Forests:**
- Complex, non-linear relationships between features
- High-dimensional datasets with many interactions
- Noisy data where ensemble methods can reduce variance
- When interpretability is less important than performance

**When Linear Regression Might Suffice:**
- Relatively simple, linear relationships
- Homogeneous datasets from similar contexts
- When interpretability is crucial
- When computational efficiency matters

This analysis demonstrates that **the best model depends on the data**. While random forests are powerful tools, they're not always necessary. The key is understanding your data and choosing the appropriate tool for the job.


